Density-dependent selection and the limits of relative fitness

### 2018-03-06 manuscript.tex
l

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures may be more useful in different circumstances. But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Constant relative fitness values can be justified as a linear approximation \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994} which is close to exact provided that selection is sufficiently weak and stable over time. Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}. The question is whether relative fitness can be used when selection is not vanishingly weak. In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap. 4]{charlesworth_1994}. We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.

The basis of relative fitness is straightforward in the absence of crowding: it simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Population genetics traditionally evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.

@@

Each iteration, adults produce new offspring (``propagules''). These disperse at random, regardless of distance from their parents, and independently of each other (e.g. there is no avoidance of territories crowded with propagules). We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed. We assume that each adult from type $i$ produces a constant number $b_i$ of successfully dispersing propagules; the number of propagules dispersing to unoccupied territories is then given by $m_i=b_in_iU/T$. The total number of these propagules will be denoted $M=\sum_i m_i$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.

The number of propagules of the $i$'th type landing in any particular territory is denoted $x_i$. We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density for type $i$ (the total propagule density will be denoted $L=\sum_i l_i$). This is strictly only an approximation of random dispersal, but it is an excellent approximation given our assumption that the $n_i$ are large enough that drift can be ignored (Appendix A).

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\cdots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
where the sum only includes territories with at least one propagule present.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of the analytical approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

To see why this connection potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.

Starting from a monomorphic population, the frequency of a $\delta_i\rightarrow \delta_i(1-\epsilon)$ variant obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of a $b_i\rightarrow b_i(1+\epsilon)$ variant will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$. The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, noticeable deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.2$ and upwards i.e. selection must be quite strong (Fig.~\ref{fig:strengthofselection}).

@@

Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above. The reason is that $b$ regulates density by controlling how many unoccupied territories receive propagules. Thus, greater $b$ means more propagules contesting territories, but also more territories being contested. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases. Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details). Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.

@@

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely; \citealt{travis_2013}). For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.

@@

\section*{Appendix A: Poisson dispersal}

For simplicity of presentation, we assume a Poisson distribution for the $x_i$ as our model of dispersal. Strictly speaking, the total number of $i$ propagules $\sum x_i$ (summed over unoccupied territories) is then no longer a constant $m_i$, but fluctuates between generations for a given mean $m_i$. Nevertheless, since we do not consider the random fluctuations in type abundances here, and for ease of comparison with the classic lottery model, we ignore the fluctuations in $m_i$. Instead we focus on Poisson fluctuations in propagule composition in each territory.

In the exact model of random dispersal, the counts of a type's propagules across unnocupied territories follows a multinomial distribution with dimension $U$, total number of trials equal to $m_i$, and equal probabilities $1/U$ for a propagule to land in a given territory. Thus, the $x_i$ in different territories are not independent random variables. However, for sufficiently large $U$ and $m_i$, this multinomial distribution for the $x_i$ across territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ \citep[Theorem 1]{arenbaev_1977}. Since we are ignoring finite population size effects, we effectively have $T\rightarrow \infty$, in which case $U$ can only be small enough to violate the Poisson approximation if there is vanishing population turnover, and then the dispersal distribution is irrelevant anyway. Likewise, in ignoring stochastic finite population size for the $n_i$, we have effectively already assumed that $m_i$ is large enough to justify the Poisson approximation (the error scales as $1/\sqrt{m_i}$; \citealt{arenbaev_1977}).

\section*{Appendix B: Growth equation derivation}

@@

\section*{Appendix C: Total density in the Lotka-Volterra competition model}

@@

\section*{Appendix D: Density-dependence of $b$-selection}

>>>>

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures may be more useful in different circumstances. But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Constant relative fitness values can be justified as an approximation which holds when selection is sufficiently weak and stable over time. Formally, these values are the ``zeroth order'' components of the actual relative fitnesses, which are not constant but depend on density, frequency, age structure, and so on \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994}. Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density also varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}. The question is whether relative fitness can be used when selection is not vanishingly weak. In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap. 4]{charlesworth_1994}. We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.

In the absence of crowding, relative fitness simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. The change in frequency depends only the relative ratio $\frac{W_i}{\overline{W}}$, not the absolute values of the growth factors $W_i$. Thus, we can replace the $W_i$ with any set of values $w_i$ that are proportional to the $W_i$; these values are called ``relative fitnesses''. The analogous formula in continuous time is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W_i$ is replaced by the intrinsic exponential growth rate $r_i$ \citep[pp. 26]{crow_1970}, and the ratio $\frac{W_i}{\overline{W}}$ is replaced by the difference $r_i-\overline{r}$. In the particular case that there are two types present, a wildtype $i$ and a mutant $j$ for instance, then the continuous time equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient.

The simple interpretation of relative fitness as a difference in intrinsic growth rates breaks down when we allow for population crowding. Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Population genetics traditionally evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.

@@

Each iteration, adults produce new offspring (``propagules''). These disperse at random, regardless of distance from their parents, and independently of each other (e.g. there is no avoidance of territories crowded with propagules). To model the dispersal process, we assume that the number of propagules of the $i$'th type landing in any particular territory, denoted $x_i$, follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$. Here $l_i=m_i/U$ is the mean territorial propagule density, where $m_i$ is the mean number of propagules produced by type $i$ (the total propagule density will be denoted by $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$). We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed. Thus, the number of propagules dispersing to unoccupied territories can be written as $m_i=b_in_iU/T$, where $b_i$ is the number of propagules produced by each adult from type $i$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\cdots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\cdots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
where the sum only includes territories with at least one propagule present.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of the analytical approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

To see why this connection potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $1/K$ in the logistic model.

Starting from a monomorphic population, the frequency of a variant with $\delta_j=\delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_j}{dt}=\epsilon \delta_i N p_j(1-p_j). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of a variant with $b_j= b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$. The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, noticeable deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.2$ and upwards i.e. selection must be quite strong (Fig.~\ref{fig:strengthofselection}).

@@

Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above. The reason is that $b$ regulates density by controlling how many unoccupied territories receive propagules. Thus, greater $b$ means more propagules contesting territories, but also more territories being contested. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases. Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_j= b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details). Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.

@@

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely; \citealt{travis_2013}). For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.

@@

\section*{Appendix A: Growth equation derivation}

@@

\section*{Appendix B: Total density in the Lotka-Volterra competition model}

@@

\section*{Appendix C: Density-dependence of $b$-selection}
